Diagnosing Child Poverty Reduction Interventions

### 2024-09-18 minos/outcomes/CPR_Intervention_Diagnostics.Rmd
Testing clipping the hh_income minimum prediction after scaling to ensure minimum is not lower than the minimum in input data. Not clipping maximum as tests werent great for that

library(tidyverse)
library(ggplot2)
library(ggridges) # for densities over time plots. 
library(knitr)
library(here)
library(gghighlight)
library(spatstat)
library(ggalluvial)
library(gridExtra)
library(circlize)

@@

```{r, include=FALSE}
out.path <- here::here('output', 'default_config/')
#out.path <- here::here('output', 'scotland_scaled/')
base.dat <- read_singular_local_out(out.path, 
                                    'baseline', 
                                    drop.dead = TRUE, 
                                    drop.zero.weight = TRUE)
int.dat <- read_singular_local_out(out.path,
                               'ChildPovertyReduction',
                               drop.dead = TRUE,
                               drop.zero.weight = TRUE)
# rel.int <- read_singular_local_out(out.path,
#                                      'ChildPovertyReductionRELATIVE_2',
#                                      drop.dead = TRUE,
#                                      drop.zero.weight = TRUE)
# abs.int <- read_singular_local_out(out.path,
#                                    'ChildPovertyReductionABSOLUTE',
#                                    drop.dead = TRUE,
#                                    drop.zero.weight = TRUE)
```

@@

ggplot(inc, aes(x = time, y = SF_12, group = scen, color = scen, fill = scen)) +
  geom_line() +
  geom_point()
```

@@

p <- density_plot(base.dat, int.dat)

p <- p + facet_wrap(~ income_quint, ncol=1)

print(p)

@@

```{r}
# Prepare data for heatmap
df_heatmap <- base.dat %>%
  filter(time %in% c(2020, 2021)) %>% # Select two time points
  select(pidp, time, income_quint) %>%
  spread(time, income_quint) %>%
  rename(Quintile_T1 = `2020`, Quintile_T2 = `2021`) %>%
  group_by(Quintile_T1, Quintile_T2) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  complete(Quintile_T1 = 1:5, Quintile_T2 = 1:5, fill = list(count = 0))

ggplot(df_heatmap, aes(Quintile_T1, Quintile_T2, fill = count)) +
  geom_tile() +
  geom_text(aes(label = count), color = "black", size = 4) +  # Add count labels in each tile
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Transition Heatmap between Quintiles",
       x = "Quintile at Baseline", y = "Quintile at Intervention") +
  theme_minimal()
```

@@

# Example usage:
plot_transition_heatmaps(base.dat, int.dat)

@@

# Chord Diagram

@@

>>>>

library(tidyverse)
library(ggplot2)
library(ggridges) # for densities over time plots. 
library(knitr)
library(here)
library(gghighlight)
library(spatstat)
library(ggalluvial)
library(gridExtra)
library(circlize)
library(viridis)

@@

```{r, include=FALSE}
#out.path <- here::here('output', 'default_config/')
out.path <- here::here('output', 'scotland_scaled/')
base.dat <- read_singular_local_out(out.path, 
                                    'baseline', 
                                    drop.dead = TRUE, 
                                    drop.zero.weight = TRUE)
int.dat <- read_singular_local_out(out.path,
                               'ChildPovertyReduction',
                               drop.dead = TRUE,
                               drop.zero.weight = TRUE)
# rel.int <- read_singular_local_out(out.path,
#                                      'ChildPovertyReductionRELATIVE_2',
#                                      drop.dead = TRUE,
#                                      drop.zero.weight = TRUE)
# abs.int <- read_singular_local_out(out.path,
#                                    'ChildPovertyReductionABSOLUTE',
#                                    drop.dead = TRUE,
#                                    drop.zero.weight = TRUE)
```

@@

rm(boosted.pidps)
```

@@

# p <- density_plot(base.dat, int.dat)
# 
# p <- p + facet_wrap(~ income_quint, ncol=1)
# 
# print(p)

rm(p)

@@

```{r}
# Prepare data for heatmap
# df_heatmap <- base.dat %>%
#   filter(time %in% c(2020, 2021)) %>% # Select two time points
#   select(pidp, time, income_quint) %>%
#   spread(time, income_quint) %>%
#   rename(Quintile_T1 = `2020`, Quintile_T2 = `2021`) %>%
#   group_by(Quintile_T1, Quintile_T2) %>%
#   summarise(count = n()) %>%
#   ungroup() %>%
#   complete(Quintile_T1 = 1:5, Quintile_T2 = 1:5, fill = list(count = 0))
# 
# 
# ggplot(df_heatmap, aes(Quintile_T1, Quintile_T2, fill = count)) +
#   geom_tile() +
#   geom_text(aes(label = count), color = "black", size = 4) +  # Add count labels in each tile
#   scale_fill_gradient(low = "white", high = "blue") +
#   labs(title = "Transition Heatmap between Quintiles",
#        x = "Quintile at Baseline", y = "Quintile at Intervention") +
#   theme_minimal()
```

@@

# Example usage:
#plot_transition_heatmaps(base.dat, int.dat)

@@

## Chord Diagram

@@

## Income Growth Trajectories

Quick bit of data prep.

```{r}
base.dat <- base.dat %>%
  select(pidp, hidp, time, income_quint, hh_income, SF_12) %>%
  mutate(scenario = 'Baseline')
int.dat <- int.dat %>%
  select(pidp, hidp, time, income_quint, hh_income, SF_12) %>%
  mutate(scenario = 'Intervention')

all.dat <- rbind(base.dat, int.dat)
```

```{r}
# Calculate average income growth by quintile and year
income_growth <- all.dat %>%
  group_by(time, income_quint, scenario) %>%
  summarise(avg_income = mean(hh_income)) %>%
  mutate(income_growth = avg_income - lag(avg_income, 1)) %>%
  filter(!is.na(income_growth))

# Plot
ggplot(income_growth, aes(x = time, y = income_growth, color = scenario)) +
  geom_line() +
  facet_wrap(~income_quint) +
  labs(title = "Income Growth by Quintile Over Time",
       x = "Year", y = "Income Growth") +
  theme_minimal()
```

## SF-12 MCS Distribution Shifts

```{r}
# Plot SF-12 MCS distribution by quintile
ggplot(all.dat, aes(x = SF_12, fill = scenario)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~income_quint) +
  labs(title = "SF-12 MCS Distribution by Quintile and Scenario",
       x = "SF-12 MCS", y = "Density") +
  theme_minimal()

# ggplot(filter(all.dat, scenario == 'Baseline'), aes(x = SF_12, fill = scenario)) +
#   geom_density(alpha = 0.5) +
#   facet_wrap(~income_quint) +
#   labs(title = "SF-12 MCS Distribution by Quintile and Scenario",
#        x = "SF-12 MCS", y = "Density") +
#   theme_minimal()
```

## Change in SF-12

```{r}
# Calculate change in SF-12 MCS by quintile and year
sf12_change <- all.dat %>%
  group_by(time, income_quint, scenario) %>%
  summarise(mean_sf12 = mean(SF_12)) %>%
  spread(scenario, mean_sf12) %>%
  mutate(sf12_diff = Intervention - Baseline)

# Plot
ggplot(sf12_change, aes(x = time, y = sf12_diff, fill = as.factor(income_quint))) +
  geom_col(position = "dodge") +
  labs(title = "Change in SF-12 MCS by Quintile",
       x = "Year", y = "SF-12 MCS Change") +
  theme_minimal()
```

## Income vs SF-12

```{r}
# Scatter plot of income change vs SF-12 MCS change by quintile
# sf12_income_corr <- all.dat %>%
#   group_by(pidp, time, income_quint, scenario) %>%
#   summarise(income_change = hh_income - lag(hh_income),
#             sf12_change = SF_12 - lag(SF_12)) %>%
#   filter(!is.na(income_change) & !is.na(sf12_change))

sf12_income_corr <- all.dat %>%
  select(pidp, time, income_quint, scenario, hh_income, SF_12) %>%
  arrange(pidp, time) %>%  # Ensure data is arranged by individual and time
  group_by(pidp, income_quint, scenario) %>%
  mutate(income_change = hh_income - lag(hh_income),  # Change in household income
         sf12_change = SF_12 - lag(SF_12)) %>%  # Change in SF-12 MCS
  filter(!is.na(income_change) & !is.na(sf12_change))  # Filter out NA changes

ggplot(sf12_income_corr, aes(x = income_change, y = sf12_change, color = scenario)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~income_quint) +
  labs(title = "Income Change vs. SF-12 MCS Change",
       x = "Income Change", y = "SF-12 MCS Change") +
  theme_minimal()
```

## Heatmap of SF-12 Transitions

```{r}
# Categorize SF-12 MCS transitions: "Improving", "Stable", "Worsening"
prepare_sf12_transitions <- function(data) {
  data %>%
    select(pidp, time, income_quint, SF12_MCS) %>%
    group_by(pidp) %>%
    arrange(time) %>%  # Ensure data is ordered by time
    mutate(SF12_MCS_change = SF12_MCS - lag(SF12_MCS)) %>%
    mutate(SF12_MCS_transition = case_when(
      SF12_MCS_change > 0 ~ "Improving",
      SF12_MCS_change == 0 ~ "Stable",
      SF12_MCS_change < 0 ~ "Worsening",
      TRUE ~ NA_character_
    )) %>%
    filter(!is.na(SF12_MCS_transition)) %>%
    ungroup()
}

# Prepare the data
sf12_transitions <- prepare_sf12_transitions(data)

# Calculate the count of transitions per quintile and transition type
sf12_transition_counts <- sf12_transitions %>%
  group_by(income_quint, SF12_MCS_transition) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  spread(SF12_MCS_transition, count, fill = 0)

# Normalize the counts to show proportions per quintile
sf12_transition_counts <- sf12_transition_counts %>%
  gather(SF12_MCS_transition, count, -income_quint) %>%
  group_by(income_quint) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

# Plot the heatmap
ggplot(sf12_transition_counts, aes(x = SF12_MCS_transition, y = as.factor(income_quint), fill = proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "SF-12 MCS Transitions Heatmap by Quintile",
       x = "SF-12 MCS Transition", y = "Income Quintile", fill = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```{r}
# Function to prepare data for heatmap of SF-12 MCS transitions
prepare_sf12_heatmap_data <- function(data) {
  data %>%
    arrange(pidp, time) %>%  # Arrange data by individual and time
    group_by(pidp) %>%
    mutate(SF12_T1 = lag(SF_12),  # Previous year's SF-12 score
           SF12_T2 = SF_12) %>%   # Current year's SF-12 score
    filter(!is.na(SF12_T1)) %>%  # Remove rows without previous year
    group_by(SF12_T1, SF12_T2) %>%
    summarise(count = n()) %>%  # Count the number of transitions
    ungroup() %>%
    complete(SF12_T1, SF12_T2, fill = list(count = 0))  # Ensure all possible combinations
}

# Function to plot heatmap showing the differences between baseline and intervention transitions
plot_sf12_heatmap_difference <- function(base.dat, int.dat) {
  # Prepare baseline and intervention data
  base_heatmap <- prepare_sf12_heatmap_data(base.dat) %>%
    rename(count_baseline = count)
  int_heatmap <- prepare_sf12_heatmap_data(int.dat) %>%
    rename(count_intervention = count)

  # Combine both datasets and calculate differences in counts and proportions
  combined_heatmap <- base_heatmap %>%
    full_join(int_heatmap, by = c("SF12_T1", "SF12_T2")) %>%
    mutate(count_baseline = coalesce(count_baseline, 0),
           count_intervention = coalesce(count_intervention, 0),
           diff_count = count_intervention - count_baseline,
           total_count = count_baseline + count_intervention,
           diff_prop = ifelse(total_count > 0, diff_count / total_count, 0)) %>%
    replace_na(list(diff_count = 0, diff_prop = 0))  # Handle missing values

  # Plot heatmap of count differences
  count_heatmap <- ggplot(combined_heatmap, aes(SF12_T1, SF12_T2, fill = diff_count)) +
    geom_tile() +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
    labs(title = "Difference in SF-12 MCS Transitions (Counts)", 
         x = "SF-12 MCS (Previous Year)", 
         y = "SF-12 MCS (Current Year)", 
         fill = "Count Difference") +
    theme_minimal()

  # Plot heatmap of proportion differences
  prop_heatmap <- ggplot(combined_heatmap, aes(SF12_T1, SF12_T2, fill = diff_prop)) +
    geom_tile() +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
    labs(title = "Difference in SF-12 MCS Transitions (Proportions)", 
         x = "SF-12 MCS (Previous Year)", 
         y = "SF-12 MCS (Current Year)", 
         fill = "Proportion Difference") +
    theme_minimal()

  # Return both plots
  list(count_heatmap = count_heatmap, prop_heatmap = prop_heatmap)
}

# Example usage
plots <- plot_sf12_heatmap_difference(base.dat, int.dat)
plots$count_heatmap  # Display the count heatmap
plots$prop_heatmap  # Display the proportion heatmap
